---
title: "Basic dataviz with R and ggplot2"
author: "a practical by [Yan Holtz](https://github.com/holtzy)"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "yan.holtz.data@gmail.com"
linkedin: "yan-holtz-2477534a"
twitter: "r_graph_gallery"
github: "holtzy"
home: "www.yan-holtz.com"
output:
  epuRate::epurate:
    toc: TRUE
    number_sections: FALSE
    code_folding: "show"
---

```{r global options, include = FALSE}
knitr::opts_chunk$set( warning=FALSE, message=FALSE)

library(rmarkdown)
library(epuRate)

# If necessary
# library(devtools)
# install_github("holtzy/epuRate")
```



<style>
.questionNumber{
  color: #69b3a2;
  border: solid;
  border-color: #69b3a2;
  padding: 3px;
  border-width: 1px;
  border-radius: 2px;
  margin-top: 200px;
}
.code-folding-btn {
  display: none;
}

</style>


<br><br>

> This practical will lead you through the basics of `ggplot2`. It is split in 5 parts, each dedicated to a big chart family: 

- correlation
- distribution
- ranking
- part of a whole
- evolution. 


# Get ready
***
This practical will require several R packages. Install the `ggplot2` package if needed and load it:
```{r, echo=TRUE}
# Install the package if needed
#install.packages("ggplot2")

# Load it
library(ggplot2)
```

<br><br><br><span class="questionNumber">Q0.1</span> - Install and load `dplyr`, a package that will be useful all along the practical.
```{r, fig.align="center", class.source="Correction",results='hide',echo=FALSE}
# Install the package if needed
#install.packages("dplyr")

# Load it
library(dplyr)
```











# 1- Correlation
***
The first part of this practical will teach you how to build [scatterplots](https://www.data-to-viz.com/graph/scatter.html) and [bubble charts](https://www.data-to-viz.com/graph/bubble.html): the 2 most common chart types to visualize correlations.

<br><br><br><span class="questionNumber">Q1.1</span> - Load the `gapminder` dataset stored in the `gapminder` package. Have a look to the 6 first rows using the `head()` function. Describe briefly what you see as comments in your script. Check how many rows are available with `nrow()`

```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
# Install the package if needed
# install.packages(...)

# Load it
library(...)

# You now have an object called gapminder. Have a look to it
head(...)

# How many rows?
nrow(...)
```

You should get something like this:

```{r, fig.align="center", class.source="Correction",results='hide',echo=FALSE}
# Install the package if needed
#install.packages("gapminder")

# Load it
library(gapminder)

# Have a look to the first rows
head(gapminder)

# How many rows?
# nrow(gapminder)
```


<br><br><br><span class="questionNumber">Q1.2</span> - How many years are available in this dataset? How many data-points for each year? Full code is provided for this question.


```{r, eval=FALSE}
# Number of different year?
gapminder %>%
  select(year) %>%
  unique() %>%
  nrow()
# or
length(unique(gapminder$year))

# Number of country available per year?
gapminder %>%
  group_by(year) %>%
  summarize( n=n() )
```



<br><br><br><span class="questionNumber">Q1.3</span> Build a [scatterplot](https://www.r-graph-gallery.com/scatterplot) showing the relationship between `gdpPercap` and `lifeExp` in 1952. Use `geom_point()`. What do you observe?

```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
# basic scatterplot
gapminder %>%
  filter(year=="1952") %>%
  ggplot( aes(x=..., y=...)) +
    geom...
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
# basic scatterplot
gapminder %>%
  filter(year=="1952") %>%
  ggplot( aes(x=gdpPercap, y=lifeExp)) +
    geom_point()
```





<br><br><br><span class="questionNumber">Q1.4</span> On the previous chart, one country is very different. Which one is it?
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
# Number of different year?
gapminder %>%
  filter(...) 
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
# Number of different year?
gapminder %>%
  filter(year=="1952" & gdpPercap>90000) 
```







<br><br><br><span class="questionNumber">Q1.5</span> Build the same chart, but get rid of this country. What trend do you observe? Does it make sense? What's missing? What could be better?
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
# basic scatterplot
gapminder %>%
  filter(...) %>%
  ggplot( ...
```

You should get a chart like this:
```{r, fig.show="asis", class.source="Correction",fig.show="hide",echo=FALSE,          fig.show="asis"}
# basic scatterplot
gapminder %>%
  filter(year=="1952" & country!="Kuwait") %>%
  ggplot( aes(x=gdpPercap, y=lifeExp)) +
    geom_point()
```



<br><br><br><span class="questionNumber">Q1.6</span> - Color dots according to their `continent`. In the `aes()` part of the code, use the `color` argument.
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
gapminder %>%
  filter(...) %>%
  ggplot( aes(..., color= ...))  +
    ...
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
gapminder %>%
  filter(year=="1952" & country!="Kuwait") %>%
  ggplot( aes(x=gdpPercap, y=lifeExp, color=continent)) +
    geom_point()
```




<br><br><br><span class="questionNumber">Q1.7</span> - Let's observe an additional variable: make the circle size proportionnal to the population (`pop`). This is done with the `size` argument of `aes()`. How do you call this kind of chart?

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
gapminder %>%
  filter(year=="1952" & country!="Kuwait") %>%
  ggplot( aes(x=gdpPercap, y=lifeExp, color=continent, size=pop)) +
    geom_point()
```




<br><br><br><span class="questionNumber">Bonus</span> You're in advance? Try to do the following:

- custom the general appearance using the `theme_ipsum` of the `hrbrthemes` library.
- add transparency to circles to limit overlapping impact, with the `alpha` argument of `aes`
- sort your data by pop size to put the small circle on top of the chart, not hidden by big bubbles
- use the `ggplotly()` function of the `plotly` package to make this chart interactive

You should get something like this chart:
```{r, class.source="Correction",fig.show="hide",echo=FALSE}
# Additionnal packages:
library(hrbrthemes) # for general style
library(plotly)     # to make the chart interactive

# Chart
p <- gapminder %>%
  filter(year=="1952" & country!="Kuwait") %>%
  arrange(desc(pop)) %>%
  ggplot( aes(x=gdpPercap, y=lifeExp, fill=continent, size=pop)) +
    geom_point(alpha=0.7, stroke="white", shape=21) +
    theme_ipsum()

# Interactive more
ggplotly(p)
```
























# 2- Distribution
***
This second part is dedicated to the visualization of distribution. It is split in 2 parts:

- Visualizing one distribution
- Comparing distribution for several groups or variables

<br><br>







## 2.1 - One distribution

The [example dataset](https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv) provides the AirBnb night prices of ~1000 appartments on the French Riviera. Data is stored on [Github](https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv) and can be loaded in R as follow:
```{r}
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv", header=TRUE)
```






<br><br><br><span class="questionNumber">Q2.1.1</span> - How many rows in the dataset? (use `nrow()`) What is the min? The max? (use `summary()`). Do you see anything weird? What kind of chart would you build to visualize this kind of data?
```{r, class.source="Correction",fig.show="hide",echo=FALSE}
nrow(data)
summary(data)
```





<br><br><br><span class="questionNumber">Q2.1.2</span> - Build a histogram of the data with `geom_histogram()`. Are you happy with the output? How can we improve it?
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
data %>%
  ggplot( aes(...)) +
    geom_histogram()
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  ggplot( aes(x=price)) +
    geom_histogram()
```





<br><br><br><span class="questionNumber">Q2.1.3</span> - Build a histogram without prices over 1500 euros. `ggplot2` displays a warning message, why? What does it mean? What's the main caveat of histograms?
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
data %>%
  fil..  %>%
  gg...
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  filter(price<1500) %>%
  ggplot( aes(x=price)) +
    geom_histogram()
```



<br><br><br><span class="questionNumber">Q2.1.4</span> - Build the histogram with different values of `binwidth`, for prices <400. What do you observe? Is it important to play with this parameter?
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
...
    geom_histogram(binwidth = ...)
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  filter(price<400) %>%
  ggplot( aes(x=price)) +
    geom_histogram(binwidth = 2)
```




<br><br><br><span class="questionNumber">Q2.1.5</span> - Use `geom_density()` to build a density chart. Use the `fill` argument to set the color. Use the `help()` function to find out what is the equivalent of `bin_width` for density chart? Check its effect using different values.

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  filter(price<1000) %>%
  ggplot( aes(x=price)) +
    geom_density(color="transparent", fill="#69b3a2", bw=5)
```



<br><br>






## 2.2 - Several distributions

<u>Dataset</u>: questions like *What probability would you assign to the phrase `Highly likely`* were asked. Answers were given in the range 0-100. It allow to understand how people perceive probability vocabulary. Data is stored on Github and can be loaded in R as follow:
```{r}
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/Teaching/master/DATA/probability.csv", header=TRUE, sep=",")
```




<br><br><br><span class="questionNumber">Q2.2.1</span> - As usual, check data main features with `nrow()`, `summary()` or any other function you think is useful.

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
# Data size?
nrow(data)

# occurence of each word:
table(data$text)
```





<br><br><br><span class="questionNumber">Q2.2.2</span> - What kind of chart would allow to compare the 8 categories?






<br><br><br><span class="questionNumber">Q2.2.3</span> - Build a basic boxplot using the default options of `geom_boxplot`()
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
ggplot(data, aes(x=..., y=..., fill=...)) +
    geom_boxplot() 
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
ggplot(data, aes(x=text, y=value, fill=text)) +
    geom_boxplot() 
```





<br><br><br><span class="questionNumber">Q2.2.4</span> - What do you observe? Can you improve this chart? What would you change? Do you remind what the different parts of the box mean?





<br><br><br><spanclass="questionNumber">Q2.2.5</span> - Apply the following modifications to the previous boxplot:

- order groups (increasing order).
- flip X and Y axis (`coord_flip()`)
- remove legend (`theme`)

```{r, fig.align="center"}
# Library forcats to reorder data
library(forcats)

# Reorder data
data %>%
  mutate(text = fct_reorder(text, value, .fun = median)) %>%
  ggplot(aes(x=text, y=value, fill=text)) +
    geom_boxplot() +
    theme(
      legend.position = "none"
    ) +
    coord_flip()
```




<br><br><br><span class="questionNumber">Q2.2.6</span> - What is the main caveat with boxplot? How can we do better?

```{r, fig.align="center"}
# Library forcats to reorder data
library(forcats)

# Reorder data
data %>%
  mutate(text = fct_reorder(text, value, .fun = median)) %>%
  ggplot(aes(x=text, y=value, fill=text)) +
    geom_boxplot() +
    geom_jitter(color="grey") +
    theme(
      legend.position = "none"
    ) +
    coord_flip()
```






<br><br><br><span class="questionNumber">Bonus</span> You're in advance? Try to do the following:

- build a violin plot with `geom_violin()`
- search the internet to build a ridgeline chart.
- find out how to add a red circle to represent the mean of each group



























# 3- Ranking
***
Let's talk about the quantity of weapons exported by the top 50 largest exporters in 2017 ([source](http://armstrade.sipri.org/armstrade/page/toplist.php)). The dataset is available on [github](https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/7_OneCatOneNum.csv). Load it in R:

```{r}
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/7_OneCatOneNum.csv", header=TRUE, sep=",")
```






<br><br><br><span class="questionNumber">Q3.1</span> - Have a quick look to the dataset. Describe it. What kind of chart can you build with this dataset? Which one would be the best in your opinion? What are the countries on top of the ranking?

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
head(data)
nrow(data)
data %>% 
  arrange(desc(Value)) %>%
  head(5)
```




<br><br><br><span class="questionNumber">Q3.2</span> - Start with a basic barplot using `geom_bar()`.
```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
data %>%
  ggplot( ... ) +
    geom_bar(stat="identity", fill="#69b3a2")
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  ggplot( aes(x=Country, y=Value) ) +
    geom_bar(stat="identity", fill="#69b3a2")
```



<br><br><br><span class="questionNumber">Q3.3</span> - Previous barplot is a bit disappointing isn't it? What can you improve?




<br><br><br><span class="questionNumber">Q3.4</span> - Here are a few hints to improve the barplot:

- use `coord_flip()` to get an horizontal version
- check the code below to reorder countries.

```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
data %>%
  # reorder data
  arrange(Value) %>%
  mutate(Country=factor(Country, Country)) %>%
  
  # Make the plot
  ...
```

You should get that kind of output:

```{r, class.source="Correction",fig.show="hide",echo=FALSE              ,fig.show="show"}
data %>%
  filter(!is.na(Value)) %>%
  arrange(Value) %>%
  mutate(Country=factor(Country, Country)) %>%
  ggplot( aes(x=Country, y=Value) ) +
    geom_bar(stat="identity", fill="#69b3a2") +
    coord_flip() +
    xlab("")
```



<br><br><br><span class="questionNumber">Q3.5</span> - A lollipop plot is used in the same conditions than a barplot. Build it with `geom_segment()` and `geom_point()`

```{r, class.source="Correction",fig.show="hide",echo=FALSE              ,fig.show="show"}
data %>%
  filter(!is.na(Value)) %>%
  arrange(Value) %>%
  mutate(Country=factor(Country, Country)) %>%
  ggplot( aes(x=Country, y=Value) ) +
    geom_segment( aes(x=Country ,xend=Country, y=0, yend=Value), color="grey") +
    geom_point(size=3, color="#69b3a2") +
    coord_flip() +
    xlab("")
```




<br><br><br><span class="questionNumber">BONUS</span> - Visit the [treemap](https://www.r-graph-gallery.com/treemap/) section of the R graph gallery. Apply this kind of chart to the weapon dataset. You should get the following chart:

```{r, class.source="Correction",fig.show="hide",echo=FALSE              ,fig.show="show"}
# Package
library(treemap)

# Plot
treemap(data,
            
            # data
            index="Country",
            vSize="Value",
            type="index",
            
            # Main
            title="",
            palette="Dark2",

            # Borders:
            border.col=c("black"),             
            border.lwds=1,                         
        
            # Labels
            fontsize.labels=0.5,
            fontcolor.labels="white",
            fontface.labels=1,            
            bg.labels=c("transparent"),              
            align.labels=c("left", "top"),                                  
            overlap.labels=0.5,
            inflate.labels=T                        # If true, labels are bigger when rectangle is bigger.

            
            )
```


<br><br><br><span class="questionNumber">BONUS</span> - Try to use `coord_polar()` to transform the previous barplot in a [circular barplot](https://www.r-graph-gallery.com/circular-barplot/).






# 4- Evolution
***
Let's consider the evolution of the [bitcoin](https://en.wikipedia.org/wiki/Bitcoin) price between April 2013 and April 2018. Data are stored on [github](https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/3_TwoNumOrdered.csv). Load the dataset using the following code:

```{r}
# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/3_TwoNumOrdered.csv", header=T)
data$date <- as.Date(data$date)
#Here is one change
```



<br><br><br><span class="questionNumber">Q4.1</span> - Build a basic line chart showing the bitcoin price evolution using `geom_line()`. 

```{r, fig.align="center", eval=FALSE, class.source="Question",echo=TRUE }
ggplot(...) +
    geom_line() 
```

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  ggplot( aes(x=date, y=value)) +
    geom_line(color="#69b3a2")
```





<br><br><br><span class="questionNumber">Q4.2</span> - Switch to an area chart using `geom_area()`. Use the `color` and `fill` argument to customize chart colors.

```{r, class.source="Correction",fig.show="hide",echo=FALSE}
data %>%
  ggplot( aes(x=date, y=value)) +
    geom_area(color="#69b3a2", fill="#69b3a2")
```




<br><br><br><span class="questionNumber">Q4.3</span> - Select the last 10 values using `tail()`. Build a connected scatterplot using `geom_point()`, `geom_line()` and `geom_area()`.

```{r, class.source="Correction",fig.show="hide",echo=FALSE, fig.height=4}
data %>%
  tail(10) %>%
  ggplot( aes(x=date, y=value)) +
    geom_area(fill="#69b3a2", alpha=0.5) +
    geom_line(color="#69b3a2") +
    geom_point()
```



<br><br><br><span class="questionNumber">Bonus</span> - Visit the [time series](https://www.r-graph-gallery.com/time-series/) section of the R graph gallery. Try to use the HTML widget called `dygraph` to build an interactive version of this lineplot. 

<br>

You should get something like this.

<br>
```{r, class.source="Correction",fig.show="hide",echo=FALSE}
# Library
library(dygraphs)
library(xts)          # To make the convertion data-frame / xts format
library(lubridate)

# Then you can create the xts format
don <- xts(x = data$value, order.by = data$date)

# graph
dygraph(don) %>%
  dyOptions(labelsUTC = TRUE, fillGraph=TRUE, fillAlpha=0.1, drawGrid = FALSE, colors="#D8AE5A") %>%
  dyRangeSelector() %>%
  dyCrosshair(direction = "vertical") %>%
  dyHighlight(highlightCircleSize = 5, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE)  %>%
  dyRoller(rollPeriod = 1)
```
